Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-09 Manuscript/manuscriptPS.tex
affiliation JM
Ca binding

\begin{abstract}
Phosphatidylserine (PS) is a negatively
charged lipid type commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS lipids can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain poorly understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of previously published experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the PS headgroup using S-DROSS solid-state NMR spectroscopy.
The augmented data set is then used
to assess the accuracy of the PS headgroup structures in, and the cation binding to, PS-containing membranes
in the most commonly used classical molecular dynamics (MD) force fields including CHARMM36, Lipid17, MacRog, Slipids, GROMOS-CKP, Berger, and variants.
We show large discrepancies between different force fields, and that none of them reproduces the NMR data within experimental accuracy.
However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
The cation binding affinity is, in line with our previous results for PC lipids, not captured correctly by any of the PS force fields.
Moreover, the simulated response of PS headgroup to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results will pave the way for development of lipid force fields that correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

  Additional simulation details are given in the Supporting Information.
  $^*$ Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.

@@

  The additional simulation details are given in the Supporting Information.

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the Supporting Information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline

@@

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

@@

The different PS MD models produced highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the Supporting Information).
As was previously observed for PC lipids \cite{botan15}, also none of the PS models produced a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
In fact, the models perform less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\citenum{botan15}),
which complicates the interpretation of structural differences between the PC and PS. However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36, and CHARMM36-UA) indeed do qualitatively replicate the larger-than-in-PC
forking of the $\alpha$-carbon (these models give forkings of $\sim\!0.01$ or smaller for PC~\cite{botan15} and between $\sim\!0.05$ and $\sim\!0.12$ for PS) that is observed in experiments
(Fig.~\ref{HGorderParameters}).
Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon order parameter
for PS ($S^\beta_{\rm CH}\approx-0.09$, see Fig.~\ref{HGorderParametersPS}) compared to PC ($S^\beta_{\rm CH}\approx-0.03$,~see Fig.~2 in Ref.~\citenum{botan15}),
%Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
in line with experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    C--H bond order parameters, $S_{\rm CH}$, of the PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS had been measured with 0.1\,M of NaCl~\cite{browning80},
    while all the other data are with counterions only.
    The data for DOPS is at 303~K and for POPS at 298~K.
    The light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on how far the simulated order parameters were from the experimental
    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
    of two hydrogens attached to the same carbon were reproduced ({\textsf{\small F}} indicates a forking problem)
    in Fig.~\ref{HGorderParametersPS}.
    The letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see section~\ref{Ranking} in the Supporting Information.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
    in lipid headgroups.
    (A) Overlaying the C$_\gamma$, C$_\beta$, and C$_\alpha$ carbons demonstrates fluctuations around the O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup.
    (B) Overlaying the C$_\beta$, C$_\alpha$, and O$_\alpha$ atoms demonstrates fluctuations around the N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
\end{figure}

Interestingly, in the three models that best fit the experimental data,
the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
the other models yield binodal distributions (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
On the other hand, the C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedral angle rotates relatively
freely between approximately 100$^{\circ}$ and 300$^{\circ}$ in the best three models (as seen also in Fig.~\ref{HGstructuresPSandPC} (B)),
while other models yield more restricted conformations (Fig. \ref{dihedralsHG}).

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
Indeed, the characteristic conformations of the PS headgroup suggested here can be useful when interpreting experiments---however,
as none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to reveal the true conformational ensemble.

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
   Variation of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
    at 298~K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which both order parameters are shifted
    such that the lower order parameter is zero without additional ions. This is to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
\end{figure*}

Calcium binding affinity to membranes containing negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations give
coherent results for POPC headgroup order parameters as a function of
CaCl$_2$ in the POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the
experimentally observed~\cite{roux90} decrease of the POPC headgroup order parameters upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The sole exception is the CHARMM36 model when paired with the NBfix
corrections for calcium~\cite{kim16,han2018graph}. For these combinations, the modulation of order parameters is underestimated, 
indicating a weaker binding affinity than in experiments.
Notably, the CHARMM36 model together with the NBfix corrections \cite{venable13,kim16} suggest similar binding affinities of
calcium and sodium for a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in the CHARMM36 model when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
in all the other tested models. This is evident in the calcium density distributions where almost all Ca$^{2+}$ ions bind to the membrane interface in all simulation models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong response to small concentrations of CaCl$_2$, which saturates below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models---including CHARMM36---which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behavior
for the two POPS $\alpha$-carbon order parameters as a function of added calcium.
For example, both order parameters decrease in Berger, increase
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibit more complex behavior.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
Therefore, we conclude that
the improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.

>>>>

\begin{abstract}
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain poorly understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameter magnitudes, $|S_\mathrm{CH}|$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring the signs of $S_\mathrm{CH}$ in the POPS headgroup using S-DROSS solid-state NMR spectrsocopy.
This data set is then used
to assess the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes
in the most commonly used classical molecular dynamics (MD) models including CHARMM36, Lipid17, Macrog, Slipids, GROMOS-CKP, Berger and variants.
We show large discrepancies between different force-fields and that none of the models reproduces the NMR data within experimental accuracy.
However, the best MD models can detect the most essential differences between PC and PS headgroup structures.
In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity. Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for the improvement of lipid force fields that correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

  Additional simulation details are given in the supplementary information.
  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.

@@

  The additional simulation details are given in the supplementary information.

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP\footnote{Force field parameters for PS lipids generated for this work, for full details see the supplementary information.} \cite{piggot12}  & 0  & 110:22 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \citenum{POPCpopsGROMOSCKPwithNa} \tabularnewline

@@

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the supplementary information (Fig. \ref{DPslices}).

@@

The different PS MD models produce highly varied headgroup and glycerol backbone  $S_{\rm CH}$ (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the supplementary information).
As was previously observed for PC lipids \cite{botan15}, also none of the PS models produces a set of  $S_{\rm CH}$ in full quantitative agreement with the experiments.
In fact, the models perform less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs.~2 and 4 in Ref.~\cite{botan15}),
which complicates the interpretation of structural differences between the PC and PS. However, concentrating on the headgroup alone, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) indeed do replicate the larger-than-in-PC
forking of the $\alpha$-carbon that is observed in experiments (Fig.~\ref{HGorderParametersPS}).
Additionally, the Slipids force field correctly produces the significantly smaller $\beta$-carbon  $S_{\rm CH}$
for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}), in line with
experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    C--H bond order parameters, $S_{\rm CH}$, of PS headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines),
    and MD simulations with different force fields (symbols).
    Experimental data for DOPS are measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data are with counterions only.
    The data for DOPS is at 303~K and for POPS at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for all simulation values (excl. MacRog~K$^+$)
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the symbol marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/structures.eps}
  \caption{\label{HGstructuresPSandPC}
    Overlayed snapshots from simulations conducted with CHARMM36---the force field producing the best agreement with experiments---demonstrate the conformational fluctuations
    of lipid headgroups.
    (A) Overlaying  C$_\gamma$-C$_\beta$-C$_\alpha$ carbons demonstrate fluctuations around O$_\gamma$-C$_\gamma$-C$_\beta$-C$_\alpha$, C$_\gamma$-C$_\beta$-C$_\alpha$-O$_\alpha$,
    and N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedrals of the PS headgroup and
    (B) overlaying C$_\beta$-C$_\alpha$-O$_\alpha$ carbonds demonstrate fluctuations around N-C$_\beta$-C$_\alpha$-O$_\alpha$ and C$_\beta$-C$_\alpha$-O$_\alpha$-P
    dihedrals of the PS and PC headgroups. The trajectory used for CHARMM36 POPC is available at Ref.~\citenum{POPCcharmm36T303K}.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
    in Figure~\ref{HGorderParametersPS}.
    The letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

Interestingly, in the three models that best fit the experimental data,
the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral angle distribution has a single peak around 120$^{\circ}$, while
other models yield binodal distributions (Fig. \ref{dihedralsHG}).
The restricted motion is also visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS}),
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
On the other hand, the C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedral angle rotates relatively
freely between approximately 100$^{\circ}$ and 300$^{\circ}$ in the best three models (as seen also in Fig.~\ref{HGstructuresPSandPC} (B)),
while other models yield more restricted conformations (Fig. \ref{dihedralsHG}).

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=\textwidth]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
   Variation of the POPC (left) and POPS (right) headgroup order parameters in a POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and from different force fields
    at 298~K (except the data for Berger are from a simulation of POPC:POPS (4:1) mixture at 310~K \cite{ollila07a,melcrova16}). 
    The $S_\mathrm{CH}$ values from systems without calcium are set at the zero point of y-axes,
    except for the $\alpha$-carbon order parameters of POPS (bottom right), for which both $S^\alpha_\mathrm{CH}$ are shifted
    such that the lower $S^\alpha_\mathrm{CH}$ is zero without additional ions. This is to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions were used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulations with added calcium, the charge was neutralized with calcium, and monovalent counterions were not present.
  }
\end{figure*}

Calcium binding affinity to membranes containing negatively charged PS lipids can be
experimentally quantified by measuring the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (see section~\ref{electrometerFORmixtures}),
where the measurement is not compromised by the dehydrated lipid--ion complexes and phase separation, and the bilayer remains uniform~\cite{feigenson86,mattai89,roux90,roux91}.
Despite the lack of an ion-free reference state
in the presence of negatively charged lipids, our simulations gave
coherent results for the POPC headgroup order parameters as a function of
CaCl$_2$ concentration in POPC:POPS (5:1) mixtures (Fig. \ref{changesWITHCaClPS}).
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimated the
experimentally observed~\cite{roux90} decrease of the POPC headgroup $S_\mathrm{CH}$ upon increasing Ca$^{2+}$ concentration (Fig. \ref{changesWITHCaClPS}),
indicating too strong calcium binding affinity.
The sole exception was CHARMM36 when paired with the NBfix
corrections for calcium~\cite{kim16,han2018graph}; for these combinations, the modulation of order parameters was underestimated, 
indicating a weaker binding affinity than in experiments.
Notably, CHARMM36 with the NBfix corrections \cite{venable13,kim16} suggested similar binding affinities for
calcium and sodium to a POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to experiments~\cite{cevc90,akutsu81,altenbach84}. This suggests that the calcium binding affinity
is underestimated in CHARMM36 when using the NBfix for calcium \cite{kim16,han2018graph}, but overestimated 
in all the other tested force fields. This is evident in the calcium density distributions, where almost all the Ca$^{2+}$ ions bind to the membrane interface in all models except CHARMM36 (Fig. \ref{CAdensPCPSmixture}).

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong response to small concentrations of CaCl$_2$, which saturates below 100~mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon $S_\mathrm{CH}$ increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon $S_\mathrm{CH}$ decreases. Moreover, a slight increase is observed in
the smaller $\alpha$-carbon $S_\mathrm{CH}$.
In simulations, all these responses were qualitatively correct only in CHARMM36.
In all force fields, even the qualitatively correct responses were much exaggerated---also in CHARMM36 that underestimated the Ca$^{2+}$ binding affinity---with
the sole exception of the larger $S^\alpha_\mathrm{CH}$, whose change CHARMM36 underestimated.
%All these changes were significantly overestimated in the tested simulation models---including CHARMM36 that underestimated the binding affinity.
Importantly, different force fields predicted qualitatively different behavior
for the two POPS $\alpha$-carbon order parameters as a function of added calcium:
Both order parameters decreased in Berger, but increased
in MacRog, whereas Lipid14/17 and CHARMM36 models exhibited more complex behaviors.
This is in contrast to the PC headgroup, where a
qualitatively correct response to bound ions is reproduced
by all simulation models, despite the significant discrepancies in the headgroup
structures observed in salt-free simulations~\cite{catte16}.
The divergent response of Berger may arise from the ring-like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
